---
title: "storm.analysis"
author: "Iurii Dziuban"
date: "August 24, 2014"
output: html_document
---
Impact of weather events on health and economic in the USA.
======================================================

Storms and other severe weather events can cause both public health and economic
problems for communities and municipalities. Many severe events can result in
fatalities, injuries, and property damage, and preventing such outcomes to the extent
possible is a key concern.

In this project it is explored different characteristics and estimates of storm on storm database.

Synopsis
========

Tornadoes are the most dangerous weather event to the population health. Next it is excessive heat. Flash floods and thunderstorm winds caused a lot of money due to damage.


Data Processing
===============

Data is from https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2
Documentation available here: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf

Reading data
```{r cache=TRUE}
storm <- read.csv(bzfile("data/repdata-data-StormData.csv.bz2"))
```

Cleaning and assigning event types
```{r}
# number of unique event types
length(unique(storm$EVTYPE))
# translate all letters to lowercase
event_types <- tolower(storm$EVTYPE)
# replace all punct. characters with a space
event_types <- gsub("[[:blank:][:punct:]+]", " ", event_types)
length(unique(event_types))
# update the data frame
storm$EVTYPE <- event_types
```

Further analysis done on this data

Most dangerous Events to Health
================================================

To find the most harmful(fatal, injuries) event types, the number
of casualties is aggregated by the event type.

```{r}
library(plyr)
casualties <- ddply(storm, .(EVTYPE), summarize,
                    fatalities = sum(FATALITIES),
                    injuries = sum(INJURIES))

# Find events that caused most death and injury
fatal_events <- head(casualties[order(casualties$fatalities, decreasing = T), ], 10)
injury_events <- head(casualties[order(casualties$injuries, decreasing = T), ], 10)
```

Events that caused most deaths:
```{r}
fatal_events[, c("EVTYPE", "fatalities")]
```

Events that caused most injuries
```{r}
injury_events[, c("EVTYPE", "injuries")]
```

Most economic difficult (property damage, crop damage) event types
==================================

To analyze the impact of weather events on the economy, available property
damage and crop damage reportings/estimates were used.
Transformation
h -> hundred, k -> thousand, m -> million, b -> billion
```{r}
exp_transform <- function(e) {
    if (e %in% c('h', 'H'))
        return(2)
    else if (e %in% c('k', 'K'))
        return(3)
    else if (e %in% c('m', 'M'))
        return(6)
    else if (e %in% c('b', 'B'))
        return(9)
    else if (!is.na(as.numeric(e))) # if a digit
        return(as.numeric(e))
    else if (e %in% c('', '-', '?', '+'))
        return(0)
    else {
        stop("Invalid exponent value.")
    }
}
```

Damage is calculated
```{r cache=TRUE}
prop_dmg_exp <- sapply(storm$PROPDMGEXP, FUN=exp_transform)
storm$prop_dmg <- storm$PROPDMG * (10 ** prop_dmg_exp)
crop_dmg_exp <- sapply(storm$CROPDMGEXP, FUN=exp_transform)
storm$crop_dmg <- storm$CROPDMG * (10 ** crop_dmg_exp)
```

Compute the economic loss by event type
and filter out events that caused no economic loss
```{r}
library(plyr)
econ_loss <- ddply(storm, .(EVTYPE), summarize,
                   prop_dmg = sum(prop_dmg),
                   crop_dmg = sum(crop_dmg))

econ_loss <- econ_loss[(econ_loss$prop_dmg > 0 | econ_loss$crop_dmg > 0), ]
prop_dmg_events <- head(econ_loss[order(econ_loss$prop_dmg, decreasing = T), ], 10)
crop_dmg_events <- head(econ_loss[order(econ_loss$crop_dmg, decreasing = T), ], 10)
```

Events that caused most property damage
```{r}
prop_dmg_events[, c("EVTYPE", "prop_dmg")]
```

Events that caused most crop damage

```{r}
crop_dmg_events[, c("EVTYPE", "crop_dmg")]
```

Results
=======

Health impact of weather events
-------------------------------

The plot shows most dangerous weather event types. (injuries, fatalities)

```{r}
library(ggplot2)
library(gridExtra)
p1 <- ggplot(data=fatal_events,
             aes(x=reorder(EVTYPE, fatalities), y=fatalities, fill=fatalities)) +
    geom_bar(stat="identity") +
    coord_flip() +
    ylab("Number of fatalities") +
    xlab("Weather event type") +
    theme(legend.position="none")

p2 <- ggplot(data=injury_events,
             aes(x=reorder(EVTYPE, injuries), y=injuries, fill=injuries)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    ylab("Number of injuries") +
    xlab("Weather event type") +
    theme(legend.position="none")

grid.arrange(p1, p2, main="The most dangerous weather events in the USA")
```

The most dangerous weather events in USA that cause death and injuries are tornadoes.
They caused more than 5000 deaths and more than 10000 injuries for 60 years.
The second and third place are excessive heat and flash floods....

Economic impact of weather events
---------------------------------

The following plot shows the most economic difficult events (property damage, crop damage)


```{r}
library(ggplot2)
library(gridExtra)
# Set the levels in order
p1 <- ggplot(data=prop_dmg_events,
             aes(x=reorder(EVTYPE, prop_dmg), y=log10(prop_dmg), fill=prop_dmg )) +
    geom_bar(stat="identity") +
    coord_flip() +
    xlab("Weather event type") +
    ylab("Property damage, dollars (log-scale)") +
    theme(legend.position="none")

p2 <- ggplot(data=crop_dmg_events,
             aes(x=reorder(EVTYPE, crop_dmg), y=crop_dmg, fill=crop_dmg)) +
    geom_bar(stat="identity") +
    coord_flip() + 
    xlab("Weather event type") +
    ylab("Crop damage, dollars") + 
    theme(legend.position="none")

grid.arrange(p1, p2, main="Weather costs to the USA economy")
```


Flash floods and thunderstorm winds cost the most
property damages. Also floods and hails cost a lot of money.